Large-scale spatio-temporal variation in vital rates and population dynamics of an alpine bird

### 2024-09-25 Manuscript_Nater_etal.qmd

---
title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
format: html
author: Chloé R. Nater, Francesco Frassinelli, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structured citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal <!--# we don't hyphenate this in the title. Not sure which is best but we should be consistent -->variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in reproducible and semi-automated ways that increase their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

For presentation and interpretation of results, we calculated additional key demographic parameters from the model posteriors. First, we calculated area- and year-specific average population densities, $meanDens_{x, t}$, by averaging $D_{juv, x, j, t} + D_{ad, x, j, t}$ over all transect lines $j$ within area $x$ in year $t$. We then proceeded to derive area- and year-specific realized population growth rates as $meanDens_{x, t+1} / meanDens_{x, t}$. Additionally, we calculated generation time for each area using two different approaches: per-generation population growth rate [@caswell2000] and elasticity of asymptotic growth rate with respect to fecundity [@brooks2001]. <!--# Do we want to say recruitement instead of fecundity? -->

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM. To that end, we built a function-based workflow (@fig-workflow) that includes a variety of options for controlling modelling decisions such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent occupancy, whether to run MCMC chains sequentially or in parallel, etc. We then set up three pipelines for executing the workflow that differ in their degree of automation and user interface to meet different needs and resource constraints. The first is a step-by-step R script to be run manually that is suitable for exploration, development, and debugging. The second is a largely automated "R targets" pipeline [@landau2021], which allows executions through a single command and maximizes efficiency by keeping track of the "up-to-date" vs. "outdated" status of different steps in the workflow. The third is a pipeline that is run directly from command line, sets up a fully reproducible environment through Nix \[nix2004\], and parallelizes the MCMC outside of R using GNUparallel [@tange2024]. This option avoids a range of issues that can arise with R's internal parallelization (e.g. processes running even when the parent R session has been restarted, hard to debug, bad error handling, etc.) and is particularly well suited for running on servers and high-performance computing clusters. For more details on pipeline implementations and options, we refer the reader to the vignettes in the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

![Simplified graphical representation of the "targets pipeline" for the multi-area modelling setup. Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow. For the full representation, download the repository and call "targets::tar_visnetwork()". The manual R and Nix/GNUparallel implementations of the workflow have the same structure and built on the same functions and relationships.](Figures/Workflow_layout.png){#fig-workflow width="424"}

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. <!--# Do we want to start a new paragraph after this sentence? -->

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival, lower recruitment, and longer generation times) were more common in the northern and mountainous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential within coordinating structured monitoring that employs common sampling protocol, training programmes, and data processing pipelines. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on fine-scale spatio-temporal variation in demography and population dynamics at across a large area. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and rugged terrain, which may hamper detectability. Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g. @bowler2020, see also next section]. Our modelling framework can be easily adapted for studying the impact of these and other variables on detectability (see below). Additionally, our results could motivate taking a closer look at monitoring challenges and potential improvements in the northern part of the country in particular. Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.

@@

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. in modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models [CARs, @verhoef2018]. Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). Estimation of latent parameters in missing areas may be possible though (Perry de Valpine, personal communication; @schaub2021 chapter 19), and this may result in a unique opportunity for making predictions of ptarmigan population trends in unmonitored areas, provided that data for a sufficient number and range of areas are available. Here, we may benefit from the fact that the line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land. Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

@@

Producing a transparent and reproducible workflow for the analysis presented here was a central objective in this study. We have done this by setting up a well documented, function-based R workflow that allows (re-)running the complete analysis from downloading the publicly available data to visualizing the results produced by the IDSM ([@fig-workflow]) and that can be implemented using pipelines that employ "R targets" [@landau2021] and Nix/GNUparallel [@nix2004; @tange2024]. Modern applied ecology needs research to be published not just as scientific papers, but as reproducible and well documented workflows [@lewis2018]. This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners [@powers2019]. This is both because of the enhanced transparency and credibility provided by openly available reproducible workflows and because of their cost-effectiveness, which allows for more sustainable use of funding in the mid- to long-term. Finally, open and reproducible workflows facilitate collaboration and inclusion of stakeholders in the research process, paving the path for the translational science that is required for society to tackle the the biodiversity crisis [@rubert2021]. It is our hope that this study can serve as an example of where to start.

>>>>

---
title: "Large-scale spatio-temporal variation in vital rates and population dynamics of an alpine bird"
format: html
author: Chloé R. Nater, Francesco Frassinelli, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structured citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in reproducible and semi-automated ways that increase their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

For presentation and interpretation of results, we calculated additional key demographic parameters from the model posteriors. First, we calculated area- and year-specific average population densities, $meanDens_{x, t}$, by averaging $D_{juv, x, j, t} + D_{ad, x, j, t}$ over all transect lines $j$ within area $x$ in year $t$. We then proceeded to derive area- and year-specific realized population growth rates as $meanDens_{x, t+1} / meanDens_{x, t}$. Additionally, we calculated generation time for each area using two different approaches: per-generation population growth rate [@caswell2000] and elasticity of asymptotic growth rate with respect to fecundity [@brooks2001]. For our particular model, the elasticity to fecundity is equivalent to the elasticity of recruitment rate.

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM. To that end, we built a function-based workflow (@fig-workflow) that includes a variety of options for controlling modelling decisions such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent occupancy, whether to run MCMC chains sequentially or in parallel, etc. We then set up three pipelines for executing the workflow that differ in their degree of automation and user interface to meet different needs and resource constraints. The first is a step-by-step R script to be run manually that is suitable for exploration, development, and debugging. The second is a largely automated "R targets" pipeline [@landau2021], which allows executions through a single command and maximizes efficiency by keeping track of the "up-to-date" vs. "outdated" status of different steps in the workflow. The third is a pipeline that is run directly from command line, sets up a fully reproducible environment through Nix \[nix2004\], and parallelizes the MCMC outside of R using GNU parallel [@tange2024]. This option avoids a range of issues that can arise with R's internal parallelization (e.g. processes running even when the parent R session has been restarted, hard to debug, bad error handling, etc.) and is particularly well suited for running on servers and high-performance computing clusters. For more details on pipeline implementations and options, we refer the reader to the vignettes in the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

![Simplified graphical representation of the "targets pipeline" for the multi-area modelling setup. Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow. For the full representation, download the repository and call "targets::tar_visnetwork()". The manual R and Nix/GNU parallel implementations of the workflow have the same structure and built on the same functions and relationships.](Figures/Workflow_layout.png){#fig-workflow width="424"}

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival, lower recruitment, and longer generation times) were more common in the northern and mountainous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential within coordinating structured monitoring that employs common sampling protocol, training programmes, and data processing pipelines. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on fine-scale spatio-temporal variation in demography and population dynamics at across a large area. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and rugged terrain, which may hamper detectability. Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g. @bowler2020, see also next section]. Our modelling framework can be easily adapted for studying the impact of these and other variables on detectability (see below). Together with ongoing efforts of increasing the number and density of transect lines in Northern Norway, this can contribute to obtaining more precise estimates of both population density and demographic rates, and would strengthen inference particularly in areas with relatively low ptarmigan population densities and less years of data.

@@

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. in modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models [CARs, @verhoef2018]. Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). Estimation of latent parameters in missing areas may be possible though (Perry de Valpine, personal communication; @schaub2021 chapter 19), and this may result in a unique opportunity for making predictions of ptarmigan population trends in unmonitored areas, provided that data for a sufficient number and range of areas are available. Here, we may benefit from the fact that the line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land. Exploring to what degree additional data from "Hønsefuglportalen" could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

@@

Producing a transparent and reproducible workflow for the analysis presented here was a central objective in this study. We have done this by setting up a well documented, function-based R workflow that allows (re-)running the complete analysis from downloading the publicly available data to visualizing the results produced by the IDSM ([@fig-workflow]) and that can be implemented using pipelines that employ "R targets" [@landau2021] and Nix/GNU parallel [@nix2004; @tange2024]. Modern applied ecology needs research to be published not just as scientific papers, but as reproducible and well documented workflows [@lewis2018]. This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners [@powers2019]. This is both because of the enhanced transparency and credibility provided by openly available reproducible workflows and because of their cost-effectiveness, which allows for more sustainable use of funding in the mid- to long-term. Finally, open and reproducible workflows facilitate collaboration and inclusion of stakeholders in the research process, paving the path for the translational science that is required for society to tackle the the biodiversity crisis [@rubert2021]. It is our hope that this study can serve as an example of where to start.
